Spring haul-out behavior of seals in the Bering and Chukchi seas

### 2023-02-02 analysis/paper/reviewer_response.md

---
editor_options: 
  markdown: 
    wrap: 72
---

We would like to thank each of the reviewers for taking their time to
read and offer suggestions for improving the quality and readability of
the manuscript. Here, we respond directly to each of the suggested edits
and comments. These responses are organized by reviewer.

## Reviewer 1 (A. Fishbach)

> In their general comments, the reviewer mentioned that they would like
> to see a more direct framing of the problem and to see a specific
> example of how the modeled availability would have an impact on
> population abundance estimates.

We address this by ...

> **Title**\
> The reviewer suggests removing *and aerial survey detectability* from
> the title of the manuscript and suggests adding the seasonal time
> period our study focused on.

We have changed the title to: Spring haul-out behavior of seals in the
Bering and Chukchi seas

> **Abstract**\
> *Can you make a statement in your abstract that use of modeled haulout
> behavior using covariates available from regional reananalysis data
> improved the confidence intervals on availability estimates and that
> this would improve aerial survey abundance estimation efforts?*

We don't feel comfortable making a precision comparison, simply because
there are too many differences between the analyses, including (1) the
data sets are different (the present study has substantially more
records), and (2) the model structures are different (the present study
has more parameters). As such, it will be difficult to ascribe a
specific reason to precision increases or decreases.

> *Please use direct language. You are not making findings on how seals
> rely on sea ice, rather you are improving methods to estimate their
> availability for improved aerial survey population estimation efforts.
> Here you are focused on modeling ice sea availability to aerial survey
> so that aerial survey counts may be appropriately interpreted. Please
> be clear that you are considering ice seal behaviors that may be
> telemtered as "dry out of water behaviors" that may be interpreted as
> behaviors that place the seal on top of the ice and available for
> observation by an aerial survey.*

The opening sentence of the abstract is a statement regarding the
reliance of ice-associated seals on sea-ice for a range of important
life history behaviors. We think this statement is important
introduction for readers who are not familiar with ice-associated seals.
We do agree with the reviewer's suggestion for more direct language and
have made edits to the beginning of the abstract with this in mind.

> **Abstract**
>
> *Did you use previously published data, or did you collect it?  Be
> specific using an active voice so that readers may understand your
> effort.*
>
> *For better flow, please indicate the behavioral data logging time
> scale here when you describe the data collection, not later when you
> speak to the analytical construct.*

We have edited this to simply state *data from satellite-linked
bio-loggers* ... we have also altered this sentence to specifically
mention the hourly timescale for the bio-logger data

> *The term "basking" implies a subset of activities that ice seals may
> engage in when hauled out on sea ice and not an activity that can be
> directly telemetered using your data logger.  Yet, what you are
> describing is "haul-out behavior".  Please use this term here.*

edited to remove use of 'basking'

> *Can you help your naive reader understand why you consider these
> covariates to be of interest and why others were not considered.*

The main body of the paper includes this discussion. Given the limited
space and purpose of the abstract we feel it best to simply list the
covariates of interest

> **Introduction**
>
> *Please use specific terms and not metaphors so that second-language
> English readers may better understand your writing.*

We have edited 'ripple effects' to 'tangible effects'

> *What about human communities that don't live in the region, but that
> send large LNG tankers and other vessel through the region?  The loss
> of sea ice has effects on those communities and activities too.*

There is a seemingly endless list of impacts of loss of sea ice to human
communities (and ecosystems) beyond the Arctic region to the rest of the
world. However, for this sentence, we wanted to keep the reader focused
on the Arctic region.

> *Please use direct language you mean biologically important
> functions?*

The term 'life history function' is a commonly used, direct term that
refers to biologically important functions. To ensure the reader is
clearly aware of which behaviors/functions we are referring to we
provide specific examples within the sentence.

> *What does "readiness" mean in the context of a marine mammal?*

This is a good catch and we have edited the sentence to, now, read:

"Knowledge about evolutionary constraints on the timing of reproductive
and molting behavior is generally lacking, so it is difficult to predict
**how or if** ice-associated seal species **might adapt** to future
changes (e.g., by adjusting pupping or molting schedules to earlier
dates or different locales)"

> *Perhaps you mean to offer a topic sentence, stating "How
> ice-assoicated seals use sea ice as a haulout platform varies betweens
> species".*

We have adopted this suggestion

> There is so much to state about ice-associated seals in the Pacific
> Arctic that this reference to Antarctic work seems out of place here.

We have moved this reference to later in the paragraph so the first
references are about Arctic work. While the example is from the
Antarctic, it is an important, early example of analysis using
bio-logger data to estimate haul-out behavior.

> *Do you mean age-class? The term life-history implies knowledge of
> previous behavioral efforts or occupancy of regions of interest.  
> Here you are specifically interested in age-class.*

We have edited to use the term 'demographic'

### 2023-02-03 analysis/paper/London_HauloutBeahvior.Rmd
additional reviewer #1 edits

```{r setup, include = FALSE}
library(targets)
library(conflicted)
library(tidyverse)
conflict_prefer("filter", "dplyr")
conflict_prefer("select", "dplyr")
library(lubridate)
conflict_prefer("month","lubridate")
library(glue)
library(scales)
library(DBI)
library(dbplyr)
library(RPostgres)
library(pingr)
library(sf)
library(flextable)
library(gt)
library(patchwork)
library(rnaturalearth)
library(rnaturalearthdata)
library(ggspatial)
library(ggsci)
library(ggrepel)
library(gghighlight)
library(ggtext)
library(glmmLDTS)
library(mgcv)
library(ragg)
library(stickylabeller)
library(MetBrewer)

@@

The global climate disruption is causing considerable reduction in Arctic sea
ice extent, volume, and seasonal presence [@kwok2018;
@meier2014; @overland2021; @wang2017]. These changes have tangible effects on
Arctic organisms, ecosystems, and the human communities who live in the region
[@huntington2020]. Such disruptions are a particular cause of concern for the
ice-associated seals that depend on spring and early summer sea ice (March-June)
in the Bering and Chukchi seas as a platform for important life history functions,
such as pupping, nursing, breeding behavior, and molting [@boveng2009;
@boveng2013a; @cameron2010; @kelly2010]. Limited data and large knowledge gaps
complicate predictions about the ultimate effects of changes in sea ice on the
behavior, health, abundance, and distribution of these seals. To date, indices
of seal health sampled during periods of declining sea ice differ regionally
[@crawford2015; @harwood2020]. Knowledge about evolutionary constraints on the
timing of reproductive and molting behavior is generally lacking, so it is
difficult to predict the readiness with which ice-associated seal species will
be able to adapt to future changes (e.g., by adjusting pupping or molting
schedules to earlier dates or different locales). This is further complicated by
the spatio-temporal variation in the phenology of these life history events
within regions and throughout their full ranges. Additionally, trends in
abundance of these species are unknown, so it is difficult to assess the effect,
if any, declines in sea ice habitat have had, or will have, on seal demography.

@@

Statutory requirements under, for instance, the Marine Mammal Protection Act
(MMPA) and the Endangered Species Act (ESA) for timely estimates of population 
abundance and trend mean improved aerial survey effort must be paired with
improved knowledge of haul-out behavior. Several studies have contributed
estimates of the distribution and abundance of ice-associated seal species in
the Arctic using aerial surveys (e.g., [@bengtson2005a], [@conn2014a], and
[@verhoef2014a]). Such abundance studies were conducted over very large areas and
estimation of absolute abundance required making inference about numerous issues
affecting the observation of seals on ice. These included availability (only
seals on ice were available to be counted), detection probability
(observers or automated detection systems may have missed some seals on ice),
species misclassification, and possible disturbance of seals by aircraft
[@conn2014a; @verhoef2014a]. Refining these inferences will improve the
accuracy of abundance estimates and, hopefully, allow credible
predictions about the effects of climate disruptions on the abundance and
distribution of Arctic seal populations.

In this study, we used 16 years of bio-logging data to investigate the haul-out
behavior of bearded, ribbon, and spotted seals in the Bering and Chukchi seas.
Our goals were threefold. First, we wished to establish baseline estimates for
the chronology of haul-out behavior in the critical spring season for each
species across different age and sex classes. Second, we sought to refine
estimates of haul-out availability corrections for aerial surveys in order to
improve estimates of seal abundance. Previously estimated availability
correction factors (e.g., [@bengtson2005a], [@conn2014a], and [@verhoef2014a])
accounted for variables such as the time-of-day and day-of-year, but did not
investigate meteorological variables that have been shown to influence haul-out
behavior of walruses [@udevitz2009]. Third, we aimed to assess the annual
variability in haul-out timing and possible linkage to changes in the extent of
seasonal sea ice between 2005 and 2020. Our work extends the scope of previous
haul-out analyses, includes the influence of meteorological variability, and
investigates the potential impact of a changing icescape on the behavior of
these species.

@@

Data used in our study came from bio-loggers deployed on bearded, ribbon, and
spotted seals in the Bering, Chukchi, and western Beaufort seas by multiple
organizations as part of collaborative investigations from 2005 through 2020.
Seals were captured using nets and bio-loggers were attached during studies
based in coastal Alaska communities or on research ships. Ship-based capture
events occurred during spring near the southern ice edge in the Bering Sea.
Land-based capture events occurred between 2005 and 2020 from May to October,
generally between the coastal communities of Scammon Bay, Alaska in the Bering
Sea, Utqiaġvik, Alaska in the Chukchi Sea, and Nuiqsut in the Beaufort Sea. We
refer readers to the primary literature for detailed capture and bio-logger
attachment methods (see Supplemental Material, S1).

@@

Haul-out behavior data from 
`r species_n %>% .$n %>% sum()` bio-loggers deployed on 
`r species_n %>% filter(species == "Bearded seal") %>% .$n` bearded, 
`r species_n %>% filter(species == "Ribbon seal") %>% .$n` ribbon, and 
`r species_n %>% filter(species == "Spotted seal") %>% .$n` spotted seals were
subset to include only records from 1 March to 15 July between 
`r min(lubridate::year(analysis_data$haulout_dt))` and 
`r max(lubridate::year(analysis_data$haulout_dt))`. Bio-loggers were of the
'SPLASH' or 'SPOT' family of tags developed by Wildlife Computers (Redmond,
Washington, USA) and either adhered to the hair on the seal or attached through the
rear flipper inter-digital webbing. The use of bio-loggers adhered to the back
or head provide some benefits over flipper mounted devices (e.g. increased
satellite transmittal rates, locations at sea) but these are lost during the
following annual molt, which, depending on deployment date, limits the duration
of haul-out data they provide. Additionally, bio-loggers attached to the head or
dorsal region are often dry while the seal is floating at the surface, inducing
a slight positive bias in the hourly percent-dry values reported by the
bio-logger. For this study, in cases where both bio-logger types were deployed,
hourly percent-dry observations from the flipper tag were preferred.

>>>>

```{r setup, include = FALSE}
library(targets)
library(conflicted)
library(tidyverse)
conflict_prefer("filter", "dplyr")
conflict_prefer("select", "dplyr")
library(lubridate)
conflict_prefer("month","lubridate")
library(glue)
library(scales)
library(DBI)
library(dbplyr)
library(RPostgres)
library(pingr)
library(sf)
library(flextable)
library(gt)
library(patchwork)
library(rnaturalearth)
library(rnaturalearthdata)
library(ggspatial)
library(ggsci)
library(ggrepel)
library(gghighlight)
library(ggtext)
library(glmmLDTS)
library(mgcv)
library(ragg)
library(MetBrewer)

@@

The global climate disruption is causing considerable reduction in Arctic sea
ice extent, volume, and seasonal presence [@kwok2018;
@meier2014; @overland2021; @wang2017]. These changes have tangible effects on
Arctic organisms, ecosystems, and the human communities who live in the region
[@huntington2020]. Such disruptions are a particular cause of concern for the
ice-associated seals that depend on spring and early summer sea ice (March-June)
in the Bering and Chukchi seas as a platform for important life history
functions, such as pupping, nursing, breeding behavior, and molting
[@boveng2009; @boveng2013a; @cameron2010; @kelly2010]. Limited data and large
knowledge gaps complicate predictions about the ultimate effects of changes in
sea ice on the behavior, health, abundance, and distribution of these seals. To
date, indices of seal health sampled during periods of declining sea ice differ
regionally [@crawford2015; @harwood2020]. Knowledge about evolutionary
constraints on the timing of reproductive and molting behavior is generally
lacking, so it is difficult to predict how or if ice-associated seal species
adapt to future changes (e.g., by adjusting pupping or molting schedules to
earlier dates or different locales). This is further complicated by the
spatio-temporal variation in the phenology of these life history events within
regions and throughout their full ranges. Additionally, trends in abundance of
these species are unknown, so it is difficult to assess the effect, if any,
declines in sea ice habitat have had, or will have, on seal demography.

Statutory requirements (e.g., United States Marine Mammal Protection Act (MMPA),
United States Endangered Species Act (ESA)) for timely estimates of population
abundance and trend of these species mean improved aerial survey effort is
needed -- and, must be paired with improved knowledge of haul-out behavior to
ensure accurate estimates. Several studies have contributed estimates of the
distribution and abundance of ice-associated seal species in the Arctic using
aerial surveys (e.g., [@bengtson2005a], [@conn2014a], and [@verhoef2014a]). Such
abundance studies were conducted over very large areas and estimation of
absolute abundance required making inference about numerous issues affecting the
observation of seals on ice. These included availability (only seals on ice were
available to be counted), detection probability (observers or automated
detection systems may have missed some seals on ice), species misclassification,
and possible disturbance of seals by aircraft [@conn2014a; @verhoef2014a].
Refining these inferences will improve the accuracy of abundance estimates and,
hopefully, allow credible predictions about the effects of climate disruptions
on the abundance and distribution of Arctic seal populations.

@@

In this study, we used 16 years of bio-logging data to investigate the haul-out
behavior of bearded, ribbon, and spotted seals in the Bering and Chukchi seas.
Our goals were threefold. First, we wished to establish baseline estimates for
the chronology of haul-out behavior in the critical spring season for each
species across different age and sex classes. Second, we sought to refine
estimates of haul-out availability corrections for aerial surveys in order to
improve estimates of seal abundance. Previously estimated availability
correction factors (e.g., [@bengtson2005a], [@conn2014a], and [@verhoef2014a])
accounted for variables such as the time-of-day and day-of-year, but did not
investigate meteorological variables that have been shown to influence haul-out
behavior of walruses [@udevitz2009]. Third, we aimed to assess the annual
variability in haul-out timing and possible linkage to changes in the extent of
seasonal sea ice between 2005 and 2020. Our work extends the scope of previous
haul-out analyses, includes the influence of meteorological variability, and
investigates the potential impact of changing sea-ice extent on the behavior of
these species.

@@

For this study we used haul-out behavior data and location estimates from
bio-loggers deployed on bearded, ribbon, and spotted seals in the Bering,
Chukchi, and western Beaufort seas by multiple organizations as part of
collaborative investigations from 2005 through 2020. Seals were captured using
nets and bio-loggers were attached during studies based in coastal Alaska
communities or on research ships. Ship-based capture events occurred during
spring near the southern ice edge in the Bering Sea between 2005 and 2018.
Land-based capture events occurred between 2005 and 2020 from May to October,
generally between the coastal communities of Scammon Bay, Alaska in the Bering
Sea, Utqiaġvik, Alaska in the Chukchi Sea, and Nuiqsut in the Beaufort Sea. We
refer readers to the primary literature for detailed capture and bio-logger
attachment methods (see Supplemental Material, S1).

@@

We subset haul-out behavior data from 
`r species_n %>% .$n %>% sum()` bio-loggers deployed on 
`r species_n %>% filter(species == "Bearded seal") %>% .$n` bearded, 
`r species_n %>% filter(species == "Ribbon seal") %>% .$n` ribbon, and 
`r species_n %>% filter(species == "Spotted seal") %>% .$n` spotted seals 
to include only records from 1 March to 15 July between 
`r min(lubridate::year(analysis_data$haulout_dt))` and 
`r max(lubridate::year(analysis_data$haulout_dt))`. Bio-loggers were of the
'SPLASH' or 'SPOT' family of tags developed by Wildlife Computers (Redmond,
Washington, USA) and we either adhered them to the hair on the seal or attached
them through the rear flipper inter-digital webbing. The use of bio-loggers
adhered to the back or head provides some benefits over flipper mounted devices
(e.g. increased satellite transmittal rates, locations at sea) but these are
lost during the following annual molt, which, depending on deployment date,
limits the duration of haul-out data they provide. Additionally, bio-loggers
attached to the head or dorsal region are often dry while the seal is floating
at the surface, inducing a slight positive bias in the hourly percent-dry values
reported by the bio-logger. For this study, in cases where both bio-logger types
were deployed, we preferred hourly percent-dry observations from the flipper tag.>
